Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19

### 2023-05-04 manuscript/manuscript.qmd
Audrey edits

---
title: "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"
short-title: "Pandemic Pass?"
subtitle: "Research note"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
abstract: |
  In an effort to combat the COVID-19 pandemic, many countries declared states of emergency and derogated (temporarily suspended) from their international human rights treaty obligations. Did states actually end up complying with their pandemic-era derogations? Did derogating and non-derogating states implement emergency measures differently, and if so, how? Using data from the Varieties of Democracy PanDem dataset and the Oxford COVID-19 Government Response Tracker, we find that states which derogated from their international human rights obligations imposed emergency measures that were temporary and non-discriminatory. On the other hand, states that did not derogate were more likely impose discriminatory measures, enact emergency measures without time limits and violate non-derogable rights. Our results support the role that flexibility mechanisms such as derogations play in international law.
date: April 20, 2023
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
keywords:
  - human rights
  - treaty derogations
  - COVID-19
  - emergency policies
bibliography: bib.bib
link-citations: true
---

@@

What is the impact of escape clauses or emergency provisions in international law on human rights? Combating a deadly virus like SARS-CoV-2 required extraordinary public health measures. In an effort to fight the COVID-19 pandemic, governments across the world imposed restrictions on the freedom of movement, association, assembly and privacy. While countries were ostensibly free to instate these policies within their own borders, the imposition of these restrictions raised important questions about states' commitments to international human rights law. Some human rights treaties recognize states' needs for emergency measures during public health concerns and contain escape clauses that establish safeguards to prevent the permanent erosion of these rights during emergencies. The temporary suspension of state compliance with its international treaty obligations occurs through a legal action known as derogation. However, the potential abuse of emergency measures during derogations raises concerns about their impact on human rights. Under emergency decrees, states often impose restrictions that are temporary in theory but difficult to roll back or contain in practice [@Stasavage2020; @LuhrmannRooney2021].

Given recent medical advancements and dwindling concerns about the pandemic, the Biden administration announced that it would end the public health emergency in the US on May 11, 2023, lifting measures instituted at the beginning of the pandemic—a move that has occurred in many other countries as well. With this move, we can assess important descriptive and analytical questions raised regarding states' respect for international law and human rights during this emergency period. What did derogations mean for compliance with international law? Did derogating and non-derogating states differ in terms of how they implemented emergency measures during COVID-19? Were derogating states more likely to exploit emergency provisions and to abuse human rights during COVID-19? These questions are important for the study of international law, compliance, and human rights during times of crises.

Typically, when international relations scholars write about human rights treaty compliance, they write of positive human rights recognition. Compliance is an adherence to the expectations and written standards within the treaties, often described as a "state of conformity or identity between an actor’s behavior and a specified rule" [@RaustialaSlaughter:2002, 539]. However, there are treaty qualifications that are legal addendums to commitment that alter or even limit compliance expectations. When states comply with these legal actions, they are in fact, limiting their obligations to human rights standards. In this research note, we focus on one such legal addendum—the treaty derogation. If and when states adhere to the written expectations of treaty derogations, they will be suspending human rights recognition during a set time of domestic emergency and crisis. If states are complying with the derogation expectations, we should in fact expect a reduction in compliance with the overall human rights treaty during that time. We explore how the potential decline in compliance may actually support human rights during the COVID-19 period since emergency mitigation strategies also require certain rules and standards to prevent the abuse of rights.

@@

This research note makes two contributions. First, while there is a significant literature on compliance with international law, especially human rights treaties [e.g. @Simmons2009; @Comstock2021], and the flexibility tools such as reservations, understandings, and declarations (RUDs) that states invoke when they join or withdraw from such treaties [@Neumayer:2007; @Helfer:2002; @Helfer:2005], we know little about states' compliance with flexibility mechanisms such as derogations. Compliance with derogations may not necessarily be the same as compliance with the treaty and this research note provides a systematic understanding of the types of states that are likely to comply with these escape clauses. The pandemic further provides unique insight into derogation behavior because numerous states derogated under the same global conditions of a public health emergency, rather than domestic conflict or natural disaster typical of treaty derogations, making it easier to compare state behavior during these derogations cross nationally. In making this comparison, this research note makes an important descriptive contribution showing how the implementation of policy measures differed across derogating and non-derogating states during the pandemic.

Second, there is a robust literature on compliance with international law and how it can lead to improvements in domestic human rights conditions [@Simmons2009; @Merry:2009]. While derogations are intended to make sure government responses are temporary, proportional and non-discriminatory, some legal scholars warn of the possibility that some treaty actions removing legal obligations may actually contribute to more violations and abuse [e.g. @Lebret:2020; @helfer2021] as part of a broader trend of targeting groups during times of crises [@Grasse2021]. We find that overall, derogating states during the pandemic do not increase abuse and violations. This finding speaks to international legal compliance scholarship and supports the idea that states are being sincere about their intentions and not, byinlarge, using actions like derogation to cover abusive behavior.

@@

Derogations in international law authorize states to temporarily suspend their international treaty commitments during times of crises, providing them with flexibility in responding to events. Times of crises can include wars, natural disasters, internal political unrest such as terrorism, and public health emergencies. Derogations afford states the opportunity to adjust to these changing conditions [@HBHelferFariss]. When states derogate from their international legal obligations, the derogation provides vital information to both international and domestic monitoring bodies, interest groups, and advocates about which rights are suspended, for how long, and the reasoning behind these suspensions. This information allows these actors—at least in principle—to challenge measures that are excessive, vague, or outlast the intended time frame of their implementation [@helfer2021].

While legal critical assessments of derogations find importance in their use [@richardson2020], others argue that, "When a state does not bother with derogation provisions in times of emergencies where human rights are suspended, then it is a situation of unbridled state power.” [@burchill2005, 97] An abuse of state power can also potentially lead to the violation of non-derogable rights or rights whose protections cannot be affected by the declaration of a de jure emergency [@Greene:2020, 36]. These include rights such as right to life, freedom from torture, freedom from slavery, freedom of thought and conscience, among others, that are viewed as essential rights that cannot be suspended even during emergencies. Research finds results that support this notion—there is no meaningful variation in states' respect for derogable or non-derogable rights during officially declared states of emergencies [@richardsclay2012; @neumayer2013]. Even states that officially derogate from their international legal obligations are more likely to violate physical integrity rights, which are non-derogable rights [@neumayer2013; @comstock2019].

These findings suggest that states may not be discerning the significance of derogable vs. non-derogable rights status during emergencies. This also raises the question of the extent to which this distinction is a valuable one during the pandemic. Most derogations to the ICCPR have been for internal conflict or natural disaster. Is a public health emergency different from other states of emergency? Overall, there is a general expectation that following derogations, we may observe some decline in respect for human rights. However, previous studies have not explored the relationship between derogations and public health emergencies.

@@

A first question of interest before we examine the potential impact of derogations is the nature of states that are more likely to submit them in the first place. Little research has focused on derogations in political science, and research that does exist has mostly focused on the determinants of submitting derogations to the United Nations (UN) rather than the effects of these derogations. This literature tells us that states that are more democratic are more likely to issue derogations and other post-commitment actions such as reservations, understandings, and declarations [@Simmons2009; @comstock2019] and that most states facing an emergency do not, in fact, issue derogations [@HBHelferFariss]. The expectation is that states with strong democratic institutions need to convince domestic audiences of the legality and legitimacy of suspending rights during emergencies. Additionally, democratic states understand international law and international legal mechanisms to be more legitimate and recognize their importance more so than other states. Less democratic states, even though at times may be more likely to violate or evade international human rights law, are less likely to see nuanced pieces of treaty law as important.

@@

In this research note, we look at states' derogations from the International Covenant on Civil and Political Rights (ICCPR) during the pandemic. The ICCPR is a foundational human rights treaty with broad human rights coverage. Exceptional circumstances brought on by the global pandemic have led to numerous, extensive derogations to the ICCPR removing state obligations to important rights areas of assembly, movement, and others. The ICCPR was adopted by the UN General Assembly on December 19, 1966 and entered into force on March 23, 1976. It was designed to recognize "the inherent dignity and of the equal and inalienable rights of all members of the human family" and outlined dozens of articles aimed at creating conditions where everyone can enjoy their civil, political, economic, social, and cultural rights. The ICCPR is one of only a few select treaties that requires states to report on the suspension or derogation of certain civil and political rights only under specific situations of "public emergency which threatens the life of the nation" [@iccpr]. The other treaties with this requirement include the European Convention Protection of Human Rights and Fundamental Freedoms and the American Convention on Human Rights. However, the ICCPR is the only one with global, rather than regional, membership and participation to track during the pandemic. As of 2023, there are 173 states party to the ICCPR. This means that most, but not all of the UN’s 193 members had the ability to submit derogations during COVID-19. Other core human rights treaties do not contain the same requirement to submit derogation notifications whenever an anticipated violation will occur.

@@

Recognizing that public health exigencies would require states to curb civil and political rights, the UN Human Rights Office of the High Commissioner (OHCHR) published guidance stating that any emergency measures must meet specific criteria of legality, necessity, proportionality, and non-discrimination. However, not every country formally derogated—a number of countries implemented emergency measures without formally derogating from ICCPR. Over 100 countries issued emergency declarations while fewer than 30 submitted COVID-19 related derogations [@icnl-covid]. These COVID-19 emergency declarations may have longer term impacts beyond temporary human rights violations [@BrownBrechenmacherCarothers:2020]. Below, we first theorize how derogations led to variation in the implementation of emergency measures during the pandemic, followed by their impact on human rights.

@@

If states determine that emergency measures are justified without regard for international law, the processes for respecting rights during crises and restoring rights at the end of the emergency may not be followed. As such, many states used the pandemic an as excuse to centralize authority and implemented measures that did not meet international legal standards. Many states imposed in discriminatory and abusive policy enforcement—they imposed restrictions on travel or movement by minority or opposition groups. For instance, in Australia, indigenous and migrant communities were disproportionately targeted in the enforcement of movement restrictions [@ICRC2021]. In Bulgaria, checkpoint controls went into effect in neighborhoods largely composed of the Roma community [@bulgaria2020]. In India, the police used a ban on public gatherings as an excuse to get rid of a months-long sit-in protesting the country's new citizenship law seen as discriminatory towards Muslims [@bbc2020].

The pandemic also led to the violation of non-derogable rights, or rights that cannot be suspended in any circumstances, in many states. For instance, Greek officials intercepted and turned back boats filled with asylum-seekers and also summarily returned asylum-seekers at the land border with Turkey. In doing so, they both violated the non-derogable principle of non-refoulement, as well as Article 7 of ICCPR is the person sent back to another state where they are in danger of ill-treatment or torture [@HathawayStevensLim:2020]. Many other states  changed rules of confinement affecting detainees, which also potentially violated non-derogable rights. For instance, in March 2020, a French government order extended the duration of pre-trial detention, raising the issue of prison overcrowding and leading to inhumane and degrading treatment [@Lebret:2020]. As such, this was not just a violation of a non-derogable right, but also raised questions regarding the proportionality of emergency measures.

We expect that states that did not formally derogate and did not notify the ICCPR of impending rights suspensions and submit derogations were less likely to take the criteria for derogations—necessity, proportionality, non-discrimination and time delimited—seriously, and were therefore more likely to impose measures that violated these criteria. If states determined that emergency measures are justified without consideration of international law, we expect that there will be more violations.

@@

While some studies on repression, human rights, and the pandemic have focused within single country cases or across a region, we examine a global database to more broadly examine the relationship between treaty derogations, human rights, and state actions during the pandemic. We explore our hypotheses using data on COVID-19 derogations, state actions, health trends, and civil society measures that we collected from several different sources (see @tbl-data-sources), including start and end dates of ICCPR derogations [@UnitedNations:2023], formal government responses from the Oxford COVID-19 Government Response Tracker [@HaleAngristGoldszmidt:2021], indicators of the severity of government responses from the Varieties of Democracy project [@vdem-v12; @vdem-pandem], and COVID case and death counts from the World Health Organization's COVID-19 dashboard [@WHO:2023]. We include data from `r info$n_countries` countries during the first 15 months (or `r info$n_weeks` weeks) of the pandemic, from March 11, 2020 to June 30, 2021. While the Oxford Government Response Tracker provides daily data on a variety of government actions, submitting formal treaty responses to the ICCPR is less likely to be as immediate of an action as implementing domestic emergency actions like stay at home orders. Accordingly, we collapse this daily data into weekly totals, resulting in a panel with `r info$n_rows` country-week observations.

@@

We examine the relationship between formal treaty derogations and different policy and human rights outcomes in two stages (see @tbl-vars). First, to test our first set of hypotheses and explore whether derogations were associated with emergency state responses in general, we look at the connection between ICCPR derogations and five different emergency public health measures recorded by the Oxford COVID-19 Government Response Tracker [@HaleAngristGoldszmidt:2021]: (1) cancelling public events, (2) imposing restrictions on gathering, (3) closing public transit, (4) restricting internal movement, and (5) limiting international travel. Each of these outcomes is recorded as an ordered categorical variable, with levels ranging from "no measures" to increasing counts and severity of emergency measures.[^severity] As we are most interested in the overall association between derogations and emergency measures, we collapse these variables into binary outcomes that indicate whether a country had no emergency measures or some emergency measures in a given week.

@@

We include several additionally variables that are associated with and potentially confound both the decision to derogate and the decision to enact specific policies or human rights-related actions. To account for the weekly and overall severity of the pandemic in each country, we control for the count of new COVID-19 cases and the count of new COVID-19 deaths, as well as the cumulative count of cases and deaths.

@@

Our data is structured as a balanced time series cross-sectional (TSCS) panel that has a natural hierarchical structure, with `r info$n_weeks` weekly observations nested within `r info$n_countries` countries. Accordingly, we model our outcome variables with multilevel or hierarchical models, with region-specific random offsets for each intercept and a fixed global year trend. When working with binary outcomes like the declaration of general emergency measures or the violation of non-derogable rights, we use logistic regression models; when working ordered categorical outcomes like the degree of discriminatory policy or abusive enforcement, use use ordered logistic regression models. We take a Bayesian approach to better model the uncertainty inherent in measuring the effect of derogations on these broad policies.

@@

Because these models provide coefficients on the logged odds scale, and because the ordered logistic models require the incorporation of coefficients for each threshold to correctly interpret, we do not discuss the results of the models using raw model estimates (though we include complete regression tables in the appendix). Instead we rely on probability-scale predictions and marginal effects whenever possible. To do so, we calculate predicted probabilities holding all explanatory variables at their typical values (median or mode). For logistic models, we estimate the probability of having a specific policy or violating a specific human right; for ordered models, we estimate the probability of each possible level of severity. To determine the effect of derogations, we calculate the contrast in predicted probabilities when derogation is set to true vs. when it is set to false (e.g., in ordered logistic models we compare the probability of major discrimination under derogation with the probability of major discrimination without derogation). We use median values from our models' posterior distributions as point estimates and provide credible intervals using the 95% highest posterior density. For hypothesis testing, we report the posterior probability that the between-level predicted differences are above or below zero.

@@

Before exploring whether COVID-induced derogations are associated with specific human rights abuses, we first explore our first set of hypotheses to examine how derogating and non-derogating states approached the implementation of their emergency public health measures. @fig-preds-mfx-policies-1 shows the results of five logistic regression models as predicted probabilities of imposing specific emergency measures over the course of the first 15 months of the pandemic. In every case, throughout the early pandemic, countries with a formal derogation in effect are far more likely to have undertook each policy, and the difference in probabilities between derogating and non-derogating states is substantial (see @fig-preds-mfx-policies-2 for the percentage-point differences between the predictions for the two groups of states).

For instance, all countries were highly likely to cancel public events in March and April 2020, with greater than a 95% chance. Over time, countries that did not derogate became less likely to cancel events, with the probability dropping to just above 90% in June 2021. Countries that derogated, however, were essentially guaranteed to regularly cancel events throughout the entire period of study, with a nearly 100% probability. The difference in predicted probabilities between the two types of countries ranged from 3 to 8 percentage points ($p(\Delta > 0) = 1$). The differences in policy implementation are starker for other types of emergency measures. Derogating states had a nearly 100% probability of imposing restrictions on gathering across all 15 months, while non-derogating states ranged between 80% and 90%, representing a 15–18 percentage point difference ($p(\Delta > 0) = 1$). Derogating states were also far more likely to close public transit, beginning with a roughly 80% probability in March 2020 and ending in June 2021 with a 60% probability. Non-derogating states followed a similar downward trend over time, but 23–25 percentage points lower than their derogating counterparts ($p(\Delta > 0) = 1$). Restrictions on internal movement were practically guaranteed at the beginning of the pandemic for derogating states, with a nearly 100% probability, dropping to roughly 70% by June 2021. Non-derogating states, on the other hand, began the pandemic with a predicted 75% probability of limiting movement, dropping to only 25% by June 2021, representing a 20–40 percentage point difference ($p(\Delta > 0) = 1$). The only policy we observed where there was no sizable difference between the two types of states is restrictions on international travel. Both derogating and non-derogating states had a nearly 100% probability of limiting international movement across all 15 months.

@@

Derogations therefore appear to help protect against limitless emergency measures and somewhat against violations of non-derogable rights, but they do not appear to do much to protect against actual observed human rights abuses like discriminatory policies and abuse enforcement of those policies. We thus find strong support only for H~2c~—that derogating states are less likely to impose measures with no time limits. We find weak support for our hypothesis that derogating states will be less likely to violated non-derogable rights (H~2b~) and no support for our hypothesis that derogations protect against discriminatory policies (H~2c~). Derogations seem to help with bureaucratic processes and might protect against severe violations of human rights, but they do not appear to influence the overall implementation of emergency measures.

>>>>

---
title: "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"
short-title: "Pandemic Pass?"
subtitle: "Research note"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
abstract: |
  This research note asks whether states issuing pandemic-era human rights treaty derogations followed through with the emergency provisions or used them to increase discriminatory policies during a time of crisis. In an effort to combat the COVID-19 pandemic, many countries declared states of emergency and derogated (temporarily suspended) from their international human rights treaty obligations. Using data from the Varieties of Democracy PanDem dataset and the Oxford COVID-19 Government Response Tracker, we find that states that derogated from their international human rights obligations imposed emergency measures that were temporary and non-discriminatory. On the other hand, states that did not derogate were more likely impose discriminatory measures, enact emergency measures without time limits and violate non-derogable rights. Our results support the role that flexibility mechanisms such as derogations play in international law. The findings contribute to earlier studies of emergency provisions in international law by applying the COVID-19 context.
date: May 1, 2023
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
keywords:
  - human rights
  - treaty derogations
  - COVID-19
  - emergency policies
bibliography: bib.bib
link-citations: true
---

@@

What is the impact of international legal emergency provisions on human rights? Combating a deadly virus like COVID-19 required extraordinary public health measures. In an effort to fight the COVID-19 pandemic, governments across the world introduced restrictions on the freedom of movement, association, and assembly. While countries were legally able to introduce the policies within their borders, the restrictions raised important questions about states’ commitments to international human rights law. Some human rights treaties recognize states’ ability to use emergency measures during times of crisis and contain "escape clauses" that safeguard the permanent erosion of human rights during and following emergencies. The temporary suspension of state compliance from international treaty obligations occurs through a legal action called derogation. However, the potential abuse of emergency measures during derogations raises concerns about their impact on human rights. Under emergency decrees, states often impose restrictions that are temporary in theory but difficult to roll back or contain in practice  [@Stasavage2020; @LuhrmannRooney2021]. Did states use pandemic era derogations how they were intended or were they used as a "pandemic pass" to justify violating human rights?

Given recent medical advancements such as distribution of vaccines and the declining concern about COVID-19 severity, the Biden administration announced that it would end the public health emergency in the US on May 11, 2023. Doing so lifts measures instituted at the beginning of the pandemic—a move that has occurred in many other countries as well. As such, it is a worthwhile time to assess important descriptive and analytical questions about states’ respect for international law and human rights during this emergency period. This research note specifically examines what derogating meant for compliance with international law during the COVID-19 pandemic. In it, we examine whether states lived up to their expected derogations and investigate the broader concern about whether states used derogations to exploit emergency provisions and to abuse human rights during COVID-19. These questions are important for the study of international law, compliance, and human rights during times of crises.

When states comply with derogations, they actually limit their obligations to human rights treaties overall. If states comply with the derogation conditions, we should in fact expect a reduction in compliance with the overall human rights treaty during that time. We explore how the potential decline in compliance may actually be a good indication that states are using treaty design as intended.

@@

This research note makes two contributions. First, while there is a significant literature on compliance with international law, especially human rights treaties [e.g. @Simmons2009; @Comstock2021], and the flexibility tools such as reservations, understandings, and declarations (RUDs) that states invoke when they join or withdraw from such treaties [@Neumayer:2007; @Helfer:2002; @Helfer:2005], we know little about states' compliance with flexibility mechanisms such as derogations. Compliance with derogations may not necessarily be the same as compliance with the treaty and this research note provides a systematic understanding adherence to derogations. The pandemic further provides unique insight into derogation behavior because numerous states derogated under the same global conditions of a public health emergency, rather than domestic conflict or natural disaster typical of treaty derogations, making it easier to compare state behavior during these derogations cross nationally. In making this comparison, this research note makes an important descriptive contribution showing how the implementation of policy measures differed across derogating and non-derogating states during the pandemic.

Second, there is a robust literature on compliance with international law and how it can lead to improvements in domestic human rights conditions [@Simmons2009; @Merry:2009]. While derogations are intended to make sure government responses are temporary, proportional and non-discriminatory, some legal scholars warn of the possibility that some treaty actions removing legal obligations may actually contribute to more violations and abuse [e.g. @Lebret:2020; @helfer2021] as part of a broader trend of targeting groups during times of crises [@Grasse2021]. We find that overall, derogating states during the pandemic did not increase abuse and violations. This finding speaks to international legal compliance scholarship and supports the idea that states are being sincere about their intentions and not, generally, using actions like derogation to cover abusive behavior.

@@

Derogations in international law authorize states to temporarily suspend their international treaty commitments during times of crises, providing them with flexibility to respond to events. Times of crises can include wars, natural disasters, internal political unrest such as terrorism, and public health emergencies. Derogations afford states the opportunity to adjust to these changing conditions [@HBHelferFariss]. When states derogate from their international legal obligations, the derogation provides vital information to both international and domestic monitoring bodies, interest groups, and advocates about which rights are suspended, for how long, and the reasoning behind these suspensions. This information allows these actors—at least in principle—to challenge measures that are excessive, vague, or outlast the intended time frame of their implementation [@helfer2021].

While legal critical assessments of derogations find importance in their use [@richardson2020], others argue that, "When a state does not bother with derogation provisions in times of emergencies where human rights are suspended, then it is a situation of unbridled state power” [@burchill2005, 97]. An abuse of state power can also potentially lead to the violation of non-derogable rights or rights whose protections cannot be affected by the declaration of a de jure emergency [@Greene:2020, 36]. These include rights such as the right to life, freedom from torture, freedom from slavery, freedom of thought and conscience, among others, that are viewed as essential rights that cannot be suspended even during emergencies. Research finds results that support this notion—there is no meaningful variation in states' respect for derogable or non-derogable rights during officially declared states of emergencies [@richardsclay2012; @neumayer2013]. Even states that officially derogate from their international legal obligations are more likely to violate physical integrity rights, which are non-derogable rights [@neumayer2013; @comstock2019].

These findings suggest that states may not be discerning the significance of derogable vs. non-derogable rights status during emergencies. This also raises the question of the extent to which this distinction is valuable during the pandemic. Most derogations to the ICCPR have been for internal conflict or natural disaster. Is a public health emergency different from other states of emergency? Overall, there is a general expectation that following derogations, we may observe some decline in respect for human rights. However, previous studies have not explored the relationship between derogations and public health emergencies.

@@

A first question of interest is the nature of states that are more likely to submit them in the first place. Little research has focused on derogations in political science, and research that does exist has mostly focused on the determinants of submitting derogations to the UN rather than the effects of these derogations. This literature tells us that states that are more democratic are more likely to issue derogations and other post-commitment actions such as reservations, understandings, and declarations [@Simmons2009; @comstock2019] and that most states facing an emergency do not, in fact, issue derogations [@HBHelferFariss]. The expectation is that states with strong democratic institutions need to convince domestic audiences of the legality and legitimacy of suspending rights during emergencies. Additionally, democratic states understand international law and international legal mechanisms to be more legitimate and recognize the importance of them more so than other states. Less democratic states, even though at times may be more likely to violate or evade international human rights law, are less likely to see nuanced pieces of treaty law as important to engage with. Scholars were split on expectations about COVID-era derogations. @Lebret:2020 argued that derogations are the best legal mechanism for states to notify the global community of their inability to comply with human rights standards during the COVID-19 pandemic. However, @richardson2020 see derogations as a less clear path for states during COVID-19, describing the actions, their conditions, and human rights as a "tangled morass" (p. 106).

@@

In this research note, we look at states' derogations from the International Covenant on Civil and Political Rights (ICCPR) during the pandemic. The ICCPR is a foundational human rights treaty with broad human rights coverage. Exceptional circumstances brought on by the global pandemic have led to numerous, extensive derogations to the ICCPR removing state obligations to important rights areas of assembly, movement, and others. The ICCPR was adopted by the UN General Assembly on December 19, 1966, and entered into force on March 23, 1976. It was designed to recognize "the inherent dignity and of the equal and inalienable rights of all members of the human family" and outlined dozens of articles aimed at creating conditions where everyone can enjoy their civil, political, economic, social, and cultural rights. The ICCPR is one of only a few select treaties that requires states to report on the suspension or derogation of certain civil and political rights only under specific situations of "public emergency which threatens the life of the nation" [@iccpr]. The other treaties with this requirement include the European Convention Protection of Human Rights and Fundamental Freedoms and the American Convention on Human Rights. However, the ICCPR is the only one with global, rather than regional, membership and participation to track during the pandemic. As of 2023, there are 173 states party to the ICCPR. This means that most, but not all the UN’s 193 members had the ability to submit derogations during COVID-19. Other core human rights treaties do not contain the same requirement to submit derogation notifications whenever an anticipated violation will occur.

@@

Recognizing that public health exigencies would require states to curb civil and political rights, the UN Human Rights Office of the High Commissioner (OHCHR) published guidance stating that any emergency measures must meet specific criteria of legality, necessity, proportionality, and non-discrimination. However, not every country formally derogated—a number of countries implemented emergency measures without formally derogating from the ICCPR. Over 100 countries issued emergency declarations while fewer than 30 submitted COVID-19 related derogations [@icnl-covid]. These COVID-19 emergency declarations may have longer term impacts beyond temporary human rights violations [@BrownBrechenmacherCarothers:2020]. Below, we first theorize how derogations led to variation in the implementation of emergency measures during the pandemic, followed by their impact on human rights.

@@

If states determine that emergency measures are justified without regard for international law, the processes for respecting rights during crises and restoring rights at the end of the emergency may not be followed. As such, many states used the pandemic an as excuse to centralize authority and implemented measures that did not meet international legal standards. Many states imposed discriminatory and abusive policy enforcement—they imposed restrictions on travel or movement by minority or opposition groups. For instance, in Australia, indigenous and migrant communities were disproportionately targeted in the enforcement of movement restrictions [@ICRC2021]. In Bulgaria, checkpoint controls went into effect in neighborhoods largely composed of the Roma community [@bulgaria2020]. In India, the police used a ban on public gatherings as an excuse to get rid of a months-long sit-in protesting the country's new citizenship law seen as discriminatory towards Muslims [@bbc2020].

The pandemic also led to the violation of non-derogable rights, or rights that cannot be suspended in any circumstances, in many states. For instance, Greek officials intercepted and turned back boats filled with asylum-seekers and summarily returned asylum-seekers at the land border with Turkey. In doing so, they both violated the non-derogable principle of non-refoulement. Additionally, Article 7 of ICCPR prohibits people being sent back to another state where they are in danger of ill-treatment or torture [@HathawayStevensLim:2020]. Many other states changed rules of confinement affecting detainees, which also potentially violated non-derogable rights. For instance, in March 2020, a French government order extended the duration of pre-trial detention, raising the issue of prison overcrowding and leading to inhumane and degrading treatment [@Lebret:2020]. As such, this was not just a violation of a non-derogable right, but also raised questions regarding the proportionality of emergency measures.

We expect that states that did not formally derogate and did not notify the ICCPR of impending rights suspensions were less likely to take the criteria for derogations—necessity, proportionality, non-discrimination and time delimited—seriously, and were therefore more likely to impose measures that violated these criteria. If states determined that emergency measures are justified without consideration of international law, we expect that there will be more violations.

@@

While some studies on repression, human rights, and the pandemic have focused within single country cases or across a region, we examine a global database to examine the relationship more broadly between treaty derogations, human rights, and state actions during the pandemic. We explore our hypotheses using data on COVID-19 derogations, state actions, health trends, and civil society measures that we collected from several different sources (see @tbl-data-sources), including start and end dates of ICCPR derogations [@UnitedNations:2023], formal government responses from the Oxford COVID-19 Government Response Tracker [@HaleAngristGoldszmidt:2021], indicators of the severity of government responses from the Varieties of Democracy project [@vdem-v12; @vdem-pandem], and COVID case and death counts from the World Health Organization's COVID-19 dashboard [@WHO:2023]. We include data from `r info$n_countries` countries during the first 15 months (or `r info$n_weeks` weeks) of the pandemic, from March 11, 2020 to June 30, 2021. While the Oxford Government Response Tracker provides daily data on a variety of government actions, submitting formal treaty responses to the ICCPR is less likely to be as immediate of an action as implementing domestic emergency actions like stay-at-home orders. Accordingly, we collapse this daily data into weekly totals, resulting in a panel with `r info$n_rows` country-week observations.

@@

We examine the relationship between formal treaty derogations and different policy and human rights outcomes in two stages (see @tbl-vars). First, to test our first set of hypotheses and explore whether derogations were associated with emergency state responses in general, we look at the connection between ICCPR derogations and five different emergency public health measures recorded by the Oxford COVID-19 Government Response Tracker [@HaleAngristGoldszmidt:2021]: (1) cancelling public events, (2) imposing restrictions on gathering, (3) closing public transit, (4) restricting internal movement, and (5) limiting international travel. Each of these outcomes is recorded as an ordered categorical variable, with levels ranging from "no measures" to increasing counts and severity of emergency measures.[^severity] As we are most interested in the overall association between derogations and emergency measures, we collapse these variables into binary outcomes that indicate whether a country had no emergency measures or some emergency measures in each week.

@@

We include several additional variables that are associated with and potentially confound both the decision to derogate and the decision to enact specific policies or human rights-related actions. To account for the weekly and overall severity of the pandemic in each country, we control for the count of new COVID-19 cases and the count of new COVID-19 deaths, as well as the cumulative count of cases and deaths.

@@

Our data is structured as a balanced time series cross-sectional (TSCS) panel that has a natural hierarchical structure, with `r info$n_weeks` weekly observations nested within `r info$n_countries` countries. Accordingly, we model our outcome variables with multilevel or hierarchical models, with region-specific random offsets for each intercept and a fixed global year trend. When working with binary outcomes like the declaration of general emergency measures or the violation of non-derogable rights, we use logistic regression models; when working ordered categorical outcomes like the degree of discriminatory policy or abusive enforcement, we use ordered logistic regression models. We take a Bayesian approach to better model and explore the uncertainty inherent in measuring the effect of derogations on these broad policies.

@@

Because these models provide coefficients on the logged odds scale, and because the ordered logistic models require the incorporation of coefficients for each threshold to correctly interpret, we do not discuss the results of the models using raw model estimates (though we include complete regression tables in the appendix). Instead, we rely on probability-scale predictions and marginal effects whenever possible. To do so, we calculate predicted probabilities holding all explanatory variables at their typical values (median or mode). For logistic models, we estimate the probability of having a specific policy or violating a specific human right; for ordered models, we estimate the probability of each possible level of severity. To determine the effect of derogations, we calculate the contrast in predicted probabilities when derogation is set to true vs. when it is set to false (e.g., in ordered logistic models we compare the probability of major discrimination under derogation with the probability of major discrimination without derogation). We use median values from our models' posterior distributions as point estimates and provide credible intervals using the 95% highest posterior density. For hypothesis testing, we report the posterior probability that the between-level predicted differences are above or below zero.

@@

We begin by exploring our first set of hypotheses to examine how derogating and non-derogating states approached the implementation of their emergency public health measures. @fig-preds-mfx-policies-1 shows the results of five logistic regression models as predicted probabilities of imposing specific emergency measures over the course of the first 15 months of the pandemic. In every case, throughout the early pandemic, countries with a formal derogation in effect are far more likely to have undertook each policy, and the difference in probabilities between derogating and non-derogating states is substantial (see @fig-preds-mfx-policies-2 for the percentage-point differences between the predictions for the two groups of states).

For instance, all countries were highly likely to cancel public events in March and April 2020, with a greater than a 95% chance. Over time, countries that did not derogate became less likely to cancel events, with the probability dropping to just above 90% in June 2021. Countries that derogated, however, were essentially guaranteed to regularly cancel events throughout the entire period of study, with a nearly 100% probability. The difference in predicted probabilities between the two types of countries ranged from 3 to 8 percentage points ($p(\Delta > 0) = 1$). The differences in policy implementation are starker for other types of emergency measures. Derogating states had a nearly 100% probability of imposing restrictions on gathering across all 15 months, while non-derogating states ranged between 80% and 90%, representing a 15–18 percentage point difference ($p(\Delta > 0) = 1$). Derogating states were also far more likely to close public transit, beginning with a roughly 80% probability in March 2020 and ending in June 2021 with a 60% probability. Non-derogating states followed a similar downward trend over time, but 23–25 percentage points lower than their derogating counterparts ($p(\Delta > 0) = 1$). Restrictions on internal movement were practically guaranteed at the beginning of the pandemic for derogating states, with a nearly 100% probability, dropping to roughly 70% by June 2021. Non-derogating states, on the other hand, began the pandemic with a predicted 75% probability of limiting movement, dropping to only 25% by June 2021, representing a 20–40 percentage point difference ($p(\Delta > 0) = 1$). The only policy we observed where there was no sizable difference between the two types of states is restrictions on international travel. Both derogating and non-derogating states had a nearly 100% probability of limiting international movement across all 15 months.

@@

Derogations therefore appear to help protect against limitless emergency measures and somewhat against violations of non-derogable rights, but they do not appear to do much to protect against actual observed human rights abuses like discriminatory policies and abuse enforcement of those policies. We thus find strong support only for H~2c~—that derogating states are less likely to impose measures with no time limits. We find weak support for our hypothesis that derogating states will be less likely to violate non-derogable rights (H~2b~) and no support for our hypothesis that derogations protect against discriminatory policies (H~2c~). Derogations seem to help with bureaucratic processes and might protect against severe violations of human rights, but they do not appear to influence the overall implementation of emergency measures.
